---
title: "Merging metadata and recount_brain"
author: 
  - name: Leonardo Collado-Torres
    affiliation:
    - &libd Lieber Institute for Brain Development, Johns Hopkins Medical Campus
    - &ccb Center for Computational Biology, Johns Hopkins University
    email: lcolladotor@gmail.com
  - name: Ashkaun Razmara
    affiliation:
    - *libd
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
---

This document contains the code that creates the `recount_brain` table by merging all the curated SRA metadata tables created by Ashkaun Razmara. 

# Setup

First, we need some packages.

```{r 'load pkgs', warning = FALSE, message = FALSE}
library('devtools')
library('recount')
library('knitcitations')
library('BiocStyle')
```

Next we setup the citation information.

```{r 'citation info', warning = FALSE, message = FALSE}
## Load knitcitations with a clean bibliography
cleanbib()
cite_options(hyperlink = 'to.doc', citation_format = 'text', style = 'html')
# Note links won't show for now due to the following issue
# https://github.com/cboettig/knitcitations/issues/63

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation('BiocStyle'),
    devtools = citation('devtools'),
    knitcitations = citation('knitcitations'),
    knitr = citation('knitr')[3],
	recount = citation('recount')[1],
    rmarkdown = citation('rmarkdown')
)

write.bibtex(bib, file = 'merging_data.bib')
```


# Read tables

This code finds all the SRA metadata tables and reads them into R. 

```{r 'read data'}
## Find table files and name them by project id
table_files <- dir('../SRA_metadata', pattern = '.csv$', full.names = TRUE)
names(table_files) <- sapply(strsplit(table_files, ', '), function(x) {
    strsplit(x[2], '-Table')[[1]][1] })

## Read the table files
table_content <- lapply(table_files, read.csv, header = TRUE,
    stringsAsFactors = FALSE, na.strings = c('NA', ''))
```

Having read the tables, we can check how frequent are some variables that we will ignore in `recount_brain`. Most of these variables are rare and just present in a handful of studies. Others like `Sequencing.Kit` are present frequently as column names but mostly made up of NAs due to how we constructed the SRA metadata tables.

```{r 'check ignored vars'}
## Define the main variables we want from all studies
main_variables <- c('Center_Name_s', 'Library_Name_s', 'AvgSpotLen_l',
    'BioSample_s', 'Experiment_s', 'MBases_l', 'MBytes_l', 'Run_s',
    'SRA_Sample_s', 'Sample_Name_s', 'BioProject_s', 'Consent_s',
    'InsertSize_l', 'Instrument_s', 'LibraryLayout_s', 'LibrarySelection_s',
    'LibrarySource_s', 'LoadDate_s', 'Platform_s', 'ReleaseDate_s',
    'SRA_Study_s', 'Assay_Type_s', 'Organism_s', 'Brain.Bank', 'Sex',
    'Age.Units', 'Age', 'Development', 'Race', 'Sample.Origin', 'Cell.Line',
    'Tissue.Site.1', 'Tissue.Site.2', 'Tissue.Site.3', 'Brodmann.Area',
    'Hemisphere', 'Disease.Status', 'Disease', 'Tumor.Type',
    'Clinical.Stage.1', 'Clinical.Stage.2', 'Pathology', 'Viability',
    'Preparation', 'PMI', 'PMI.Units', 'RIN')

## Check whether the variables are part of our set or extra variables
## that we will ignore.
check_vars <- lapply(table_content , function(tab) {
    present <- tolower(colnames(tab))[tolower(colnames(tab)) %in%
        tolower(main_variables)]
    tobeignored <- tolower(colnames(tab))[!tolower(colnames(tab)) %in%
        tolower(main_variables)]
    return(list(present = present, tobeignored = tobeignored))
})
## Variables that will be ignored
sort(table(unlist(sapply(check_vars, '[[', 'tobeignored'))), decreasing = TRUE)

## Check that we are not missing any due to spelling issues
stopifnot(all(
    !names(table(unlist(sapply(check_vars, '[[', 'tobeignored')))) %in%
    tolower(main_variables)))
```

Next we can keep just the main variables of interest.

```{r 'keep main vars'}
## Keep only the main variables
table_content <- lapply(table_content , function(tab) {
    tab[, colnames(tab) %in% main_variables]
})
```

We can then compute some summary statistics per variable for each study, like what is the percent of missing observations.

```{r 'explore main vars'}
## Compute some statistics per variable and study
table_cols <- do.call(rbind, lapply(seq_len(length(table_content)), function(i) {
  tab <- table_content[[i]]
  df <- data.frame(columns = colnames(tab), number_na = sapply(tab, function(var) { sum(is.na(var))}), n = nrow(tab), study = names(table_content)[i], stringsAsFactors = FALSE)
  df$percent_na <- df$number_na / df$n * 100
  df$number_obs <- df$n - df$number_na
  df$percent_obs <- 100 - df$percent_na
  rownames(df) <- NULL
  return(df)
}))

## Explore the summary statistics
dim(table_cols)
summary(table_cols)

## Most commonly observed variables accross all studies
sort(table(table_cols$columns), decreasing = TRUE)
## now in percent
round(sort(table(table_cols$columns), decreasing = TRUE) /
    length(table_content) * 100, 2)

## Example of how we can explore the percent NA for the Age variable
#boxplot(percent_na ~ columns, data = table_cols, las = 2)
#hist(subset(table_cols, columns == 'Age')$percent_na)
table(subset(table_cols, columns == 'Age')$percent_na)
#summary(subset(table_cols, columns == 'Age')$percent_na)
```

# Merging tables

Having read and filtered the tables, we can now proceed to merging them. This is how we create the `recount_brain` table.

```{r 'merge tables'}
## Get the unique columns
unique_cols <- unique(table_cols$columns)
## Sort by percent NA
unique_cols_sorted <- tolower(names(sort(tapply(table_cols$percent_na,
    table_cols$columns, mean))))

## Build a new set of tables by study with all the columns
table_new <- lapply(table_content, function(tab) {
    colnames(tab) <- tolower(colnames(tab))
    missing_cols <- unique_cols[!unique_cols %in% colnames(tab)]
    if(length(missing_cols) > 0) {
        df_add <- as.data.frame(matrix(NA, nrow = nrow(tab), ncol = length(missing_cols)))
        colnames(df_add) <- missing_cols
        df_new <- cbind(tab, df_add)
        colnames(df_new) <- tolower(colnames(df_new))
        res <- df_new[, match(unique_cols_sorted, colnames(df_new))]
    } else {
        res <- tab
    }
    return(res)
})

## Finally merge them all
recount_brain <- do.call(rbind, table_new)
```

# Merge with recount

Now that we have the `recount_brain` table, we can now check if the samples are present in `recount` `r citep(bib[['recount']])` and save the data which can then be accessed easily thanks to `recount::add_metadata('recount_brain')`.

```{r 'add recount info and save'}
## Check which samples are in recount
m <- all_metadata('sra')

map <- match(m$run, recount_brain$run_s)
table(!is.na(map))

map_r <- match(recount_brain$run_s, m$run)
table(!is.na(map_r))

## Add whether the sample is present in recount
recount_brain$present_in_recount <- !is.na(map_r)

## Sort alphabetically and change dots for underscores
#recount_brain <- recount_brain[, sort(colnames(recount_brain))]
colnames(recount_brain) <- gsub('\\.', '_', colnames(recount_brain))

## Fix some issues
recount_brain$sex <- tolower(recount_brain$sex)

for(i in which(sapply(recount_brain, class) == 'character')) {
    if(any(grepl('[[:space:]]+$', recount_brain[, i]))) {
        recount_brain[, i] <- gsub('[[:space:]]+$', '', recount_brain[, i])
    }
}

## Final dimensions of recount_brain
dim(recount_brain)

## Save the data
save(recount_brain, file = 'recount_brain_v1.Rdata')
write.csv(recount_brain, file = 'recount_brain_v1.csv', quote = TRUE, row.names = FALSE)

## Check md5sum for the resulting files
sapply(dir(pattern = 'recount_brain_v1'), tools::md5sum)

## List of all variables
colnames(recount_brain)
```

# recount vs recount_brain

Not all the samples in `recount_brain` are present in `recount`. The following code explores why some samples are missing in `recount`.


```{r 'compare vs recount'}
## Number missing/present in recount
table(is.na(map_r))

## For the missing ones, what is their organism?
table(recount_brain$organism_s[ is.na(map_r) ], useNA = 'ifany')

## For the missing ones, check organism and platform
table(recount_brain$organism_s[ is.na(map_r) ],
    recount_brain$platform_s[ is.na(map_r) ], useNA = 'ifany')

## For the missing ones, check sra_study and platform
table(recount_brain$sra_study_s[ is.na(map_r) ],
    recount_brain$platform_s[ is.na(map_r) ], useNA = 'ifany')

## For all the ones in recount_brain, check assay type and assay_type_s
table(recount_brain$assay_type_s, recount_brain$assay_type, useNA = 'ifany')

## For the missing ones, check assay_type_s
table(recount_brain$assay_type_s[ is.na(map_r) ], useNA = 'ifany')
```

# Brief exploration of `recount_brain`

This section shows examples of how one can explore the data in `recount_brain`. It's up to the user to keep exploring the samples to identify questions of their interest and/or relevant studies.

```{r 'explore recount_brain 1'}
## Number of samples by disease
table(recount_brain$disease, useNA = 'ifany')

## Number of samples per age unit
table(recount_brain$age_units, useNA = 'ifany')

## Development stage vs presence in recount
table('Development stage' = recount_brain$development,
 'Present in recount' = recount_brain$present_in_recount, useNA = 'ifany')

## Age by age units for all samples in recount_brain and then those
## also present in recount
par(mar = c(10, 4, 4, 2) + 0.1)
boxplot(age ~ age_units, data = recount_brain, las = 2,
    main = 'all recount_brain samples')

boxplot(age ~ age_units, main = 'only samples present in recount',
    data = recount_brain[recount_brain$present_in_recount, ], las = 2)
    
## Get the number of samples for each boxplot
table(recount_brain$age_units[recount_brain$present_in_recount])
```

## PMI

Here is an example where we explore how many samples have the post mortem interval (PMI) information registered and the relationship with age for those that have age measured in years.

```{r 'explore recount_brain 2'}
## How many samples have a pmi unit?
table('PMI units' = recount_brain$pmi_units,
    'Present in recount' = recount_brain$present_in_recount, useNA = 'ifany')
## How many samples have a pmi value?
table(!is.na(recount_brain$pmi))

## Check PMI overall
boxplot(recount_brain$pmi, ylab = 'PMI (hours)')

## Compare PMI vs age (in years) for the samples present in recount
plot(age ~ pmi, data = recount_brain[recount_brain$present_in_recount &
    recount_brain$age_units == 'Years', ], ylab = 'Age (years)',
    xlab = 'PMI (hours)')
```

## Sex and other variables

Next we can explore the relationship between sex and other variables such as age and disease status.

```{r 'explore recount_brain 3'}
## Number of observations with sex recorded
table(recount_brain$sex[recount_brain$present_in_recount], useNA = 'ifany')

## Age vs sex for those that have age measured in years
boxplot(age ~ sex, data = recount_brain[recount_brain$present_in_recount &
    recount_brain$age_units == 'Years', ], ylab = 'Age (years)')

## Check age vs disease status
boxplot(age ~ disease_status,
    data = recount_brain[recount_brain$present_in_recount &
    recount_brain$age_units == 'Years', ], ylab = 'Age (years)')
    
## Check age disease status and sex
boxplot(age ~ disease_status + sex,
    data = recount_brain[recount_brain$present_in_recount &
    recount_brain$age_units == 'Years', ], ylab = 'Age (years)')
```

```{r 'explore recount_brain 4', fig.height = 10}
## Age vs the actual disease
par(mar = c(18, 4, 4, 2) + 0.1)
boxplot(age ~ disease, data = recount_brain[recount_brain$present_in_recount &
    recount_brain$age_units == 'Years', ], las = 2, ylab = 'Age (years)')

table(recount_brain$disease[recount_brain$present_in_recount &
    recount_brain$age_units == 'Years'], useNA = 'ifany')
```



# Reproducibility

This document was made possible thanks to:

* R `r citep(bib[['R']])`
* `r Biocpkg('BiocStyle')` `r citep(bib[['BiocStyle']])`
* `r CRANpkg('devtools')` `r citep(bib[['devtools']])`
* `r CRANpkg('knitcitations')` `r citep(bib[['knitcitations']])`
* `r CRANpkg('knitr')` `r citep(bib[['knitr']])`
* `r Biocpkg('recount')` `r citep(bib[['recount']])`
* `r CRANpkg('rmarkdown')` `r citep(bib[['rmarkdown']])`

Code for creating this document

```{r createVignette, eval=FALSE}
## Create the vignette
library('rmarkdown')
system.time(render('merging_data.Rmd', 'BiocStyle::html_document'))
```

Reproducibility information for this document.

```{r 'reproducibility info'}
## Reproducibility info
proc.time()
message(Sys.time())
options(width = 120)
session_info()
```

# Bibliography

This document was generated using `r Biocpkg('BiocStyle')` `r citep(bib[['BiocStyle']])`
with `r CRANpkg('knitr')` `r citep(bib[['knitr']])` and `r CRANpkg('rmarkdown')` `r citep(bib[['rmarkdown']])` running behind the scenes.

Citations made with `r CRANpkg('knitcitations')` `r citep(bib[['knitcitations']])` and the bibliographical file is available [here](merging_data.bib).

```{r biblio, results = 'asis', echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
bibliography()
```
