---
title: "Merging recount_brain metadata with GTEx and TCGA metadata"
author: 
  - name: Dustin Sokolowski
    affiliation:
    - &ggb Genetics and Genome Biology, Hospital for Sick Children, Toronto M5G 0A4, Canada
    - &dmg Department of Molecular Genetics, University of Toronto, Toronto M5S 1A8, Canada
    email: djsokolowski95@gmail.com
  - name: Michael D Wilson
    affiliation:
    - *ggb
    - *dmg
    - Heart and Stroke Richard Lewar Centre of Excellence in Cardiovascular Research, Toronto M5S 3H2, Canada
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
---

This document contains the code that creates the `recount_brain` version 2 table by merging `recount-brain` version 1 with the metadata in GTEx and TCGA.

# Setup

First, we need some packages.

```{r 'load pkgs', warning = FALSE, message = FALSE}
library('devtools')
library('recount')
library('knitcitations')
library('BiocStyle')
```

Next we setup the citation information.

```{r 'citation info', warning = FALSE, message = FALSE}
## Load knitcitations with a clean bibliography
cleanbib()
cite_options(hyperlink = 'to.doc', citation_format = 'text', style = 'html')
# Note links won't show for now due to the following issue
# https://github.com/cboettig/knitcitations/issues/63

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation('BiocStyle'),
    devtools = citation('devtools'),
    knitcitations = citation('knitcitations'),
    knitr = citation('knitr')[3],
	recount = citation('recount')[1],
    rmarkdown = citation('rmarkdown')
)

write.bibtex(bib, file = 'cross_studies_metadata.bib')
```


# Merging with `recount-brain`


Create a data.frame that has all the new columns (from `Sokolowski recount_brain TCGA_brain GTEx_brain index.csv`) as well as the sample id column for matching later. Include all of recount-brain v1 (all 4,431 rows) as well as all the GTEx and TCGA brain samples. Ideally, if we find a typo or something, we should be able to update the v1 table (see `merged_metadata` folder) and then the v2 table by running this Rmd. Include external files used in this folder (even if we can also download them automatically) so that we can re-run this code anywhere.

We'll also need to include `Sokolowski recount_brain TCGA_brain GTEx_brain index.csv` in the paper as a supplementary table. If you can clean it a bit that'd be great.


```{r 'create_recount_brain_v2'}
recount_brain <- add_metadata(source = 'recount_brain_v1')

## your magic here =)

## Final dimensions of recount_brain
dim(recount_brain)

## Save the data
save(recount_brain, file = 'recount_brain_v2.Rdata')
write.csv(recount_brain, file = 'recount_brain_v2.csv', quote = TRUE, row.names = FALSE)

## Check md5sum for the resulting files
sapply(dir(pattern = 'recount_brain_v2'), tools::md5sum)

## List of all variables
colnames(recount_brain)
```


# Exploring `recount-brain` v2

Add some code for exploring the new table a little bit. Some of these numbers will be mentioned in the paper.


# Reproducibility

This document was made possible thanks to:

* R `r citep(bib[['R']])`
* `r Biocpkg('BiocStyle')` `r citep(bib[['BiocStyle']])`
* `r CRANpkg('devtools')` `r citep(bib[['devtools']])`
* `r CRANpkg('knitcitations')` `r citep(bib[['knitcitations']])`
* `r CRANpkg('knitr')` `r citep(bib[['knitr']])`
* `r Biocpkg('recount')` `r citep(bib[['recount']])`
* `r CRANpkg('rmarkdown')` `r citep(bib[['rmarkdown']])`

Code for creating this document

```{r createVignette, eval=FALSE}
## Create the vignette
library('rmarkdown')
system.time(render('cross_studies_metadata.Rmd', 'BiocStyle::html_document'))
```

Reproducibility information for this document.

```{r 'reproducibility info'}
## Reproducibility info
proc.time()
message(Sys.time())
options(width = 120)
session_info()
```

# Bibliography

This document was generated using `r Biocpkg('BiocStyle')` `r citep(bib[['BiocStyle']])`
with `r CRANpkg('knitr')` `r citep(bib[['knitr']])` and `r CRANpkg('rmarkdown')` `r citep(bib[['rmarkdown']])` running behind the scenes.

Citations made with `r CRANpkg('knitcitations')` `r citep(bib[['knitcitations']])` and the bibliographical file is available [here](cross_studies_metadata.bib).

```{r biblio, results = 'asis', echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
bibliography()
```
