---
title: "Merging recount_brain metadata with GTEx and TCGA metadata"
author: 
  - name: Dustin Sokolowski
    affiliation:
    - &ggb Program in Genetics and Genome Biology, Hospital for Sick Children, Toronto M5G 0A4, Canada
    - &dmg Department of Molecular Genetics, University of Toronto, Toronto M5S 1A8, Canada
    email: djsokolowski95@gmail.com
  - name: Leonardo Collado-Torres
    affiliation:
    - &libd Lieber Institute for Brain Development, Johns Hopkins Medical Campus
    - &ccb Center for Computational Biology, Johns Hopkins University
    email: lcolladotor@gmail.com
  - name: Michael D Wilson
    affiliation:
    - *ggb
    - *dmg
    - Heart and Stroke Richard Lewar Centre of Excellence in Cardiovascular Research, Toronto M5S 3H2, Canada
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
---

This document contains the code that creates the `recount_brain` version 2 table by merging `recount-brain` version 1 with the metadata in GTEx and TCGA. Dustin Sokolowski created the `recount_brain_v2` analysis with supervision from Michael D Wilson. Leonardo Collado-Torres edited this document.



# Load packages and files. 

Here, there are two categories of files being loaded. Firstly, the `recount_brain`, TCGA, and GTEx data from `recount` are being downloaded using the `add_metadata()` and `all_metadata()` functions respectively. Secondly, some additional information about GTEx samples are added. Specifically, sample age, sex, and Hardy-Death classification are taken from the `gtex_pheno.csv`. Information in regards to to sample fixing and sample freezing are found in `gtex_sampinfo.csv`. These two files can be downloaded from https://github.com/LieberInstitute/recount-brain/tree/master/cross_studies_metadata/GTEx_extra. Furthermore, they are csv files adapted from links documented in the code.

```{r 'load pkgs', warning = FALSE, message = FALSE}
library('recount')
```

```{r 'citation info', warning = FALSE, message = FALSE, echo = FALSE}
library('devtools')
library('knitcitations')
library('BiocStyle')


## Load knitcitations with a clean bibliography
cleanbib()
cite_options(hyperlink = 'to.doc', citation_format = 'text', style = 'html')
# Note links won't show for now due to the following issue
# https://github.com/cboettig/knitcitations/issues/63
 ## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation('BiocStyle'),
    devtools = citation('devtools'),
    knitcitations = citation('knitcitations'),
    knitr = citation('knitr')[3],
	recount = citation('recount')[1],
    rmarkdown = citation('rmarkdown')
)
write.bibtex(bib, file = 'cross_studies_metadata.bib')
```

```{r}
# below are the files required to combine datasets

#GTEx & TCGA metadata from recount
recount_brain <- add_metadata(source = "recount_brain_v1")
GTEx <-  recount::all_metadata("gtex")
tcga <- recount::all_metadata("tcga")
# Read txt file downloaded from:
# "https://storage.googleapis.com/gtex_analysis_v7/annotations/GTEx_v7_Annotations_SubjectPhenotypesDS.txt"
# this dataset was also converted into a csv before being loaded into R

gtex_pheno <- read.csv("https://raw.githubusercontent.com/LieberInstitute/recount-brain/master/cross_studies_metadata/GTEx_extra/gtex_pheno.csv",
    header = T, as.is = T)

# Remaining phenotype information that may be useful for GTEx metadata
# https://storage.googleapis.com/gtex_analysis_v7/annotations/GTEx_v7_Annotations_SampleAttributesDS.txt
# This dataset was converted into a csv before being loaded into R
gtexSampinfo <- read.csv("https://raw.githubusercontent.com/LieberInstitute/recount-brain/master/cross_studies_metadata/GTEx_extra/gtex_sampinfo.csv",
    header = T, as.is = T)

# Supplementary table 1 to re-order columns
notes <- read.csv("https://raw.githubusercontent.com/LieberInstitute/recount-brain/master/SupplementaryTable1.csv",
    header= T, as.is = T)

# Generic funciton to convert factor into character vector
tochr <- function(x) return(as.character(levels(x))[x]) 
```

# Process GTEx metadata

The code chunk below extracts brain samples from GTEx using the `sms` column. Secondly, the sample id's are adjusted such that GTEx metadata from recount can be easily merged with GTEx metadata from the phenotype and sample file. Finally, these files are merged.

```{r}


GTEX_brain <- GTEx[GTEx$smts == "Brain",] # brain samples in gtex

# Change the sample id of the "sampid" column to the first 9 or 10 characters so that the GTEX_brain and gtex_pheno columns can be merged
s <- substr(GTEX_brain$sampid, 1,10) 
s1 <- c()
for(i in s) {
  last <- substr(i, nchar(i), nchar(i))
  if(last == "-") {
    s1 <- c(s1, substr(i, 1, nchar(i)-1))
  } else {
    s1 <- c(s1, i)
  }
}
GTEX_brain$SUBJID <- s1

# merge GTEX_brain and gtex_pheno
GTEx_brain_merge <- merge(GTEX_brain, gtex_pheno, by = "SUBJID")

#GTEx brain samples
gtexSampinfo_brain <- intersect(GTEx_brain_merge$sampid, gtexSampinfo$SAMPID)

```

## Tissue location

The code below processes the columns important to tissue location. Specifically, tissue location is found from `smtsd`, and locations are extracted from the same column. Finally, GTEx contains samples within the putamen, which is in the right hemisphere. Other sample regions are bilateral. 

```{r}


#Tissue location
tissue_1_gtex <- substr(GTEx_brain_merge$smtsd, 9,
    nchar(GTEx_brain_merge$smtsd))

# broadman locations
broadman_gtex <- c()
for(i in 1:nrow(GTEx_brain_merge)) {
  if(GTEx_brain_merge$smtsd[i] %in% "Brain - Anterior cingulate cortex (BA24)") {
    broadman_gtex[i] <- 24
    next
  }
  if(GTEx_brain_merge$smtsd[i] %in% "Brain - Frontal Cortex (BA9)") {
    broadman_gtex[i] <- 9
    next
  } 
  broadman_gtex[i] <- NA
}

# mapping putamen to right hemisphere
hemisphere_gtex <- c()
for(i in GTEx_brain_merge$smtsd) {
  if(i == "Brain - Putamen (basal ganglia)") {
    hemisphere_gtex <- c(hemisphere_gtex, "right")
  } else {
    hemisphere_gtex <- c(hemisphere_gtex, "bilateral")
  }
}
```

## Age, sex, disease

The code below looks at age, sex, and disease. In terms of age, every brain sample is older than 20, therefore development is only adult. Disease is organized by the Hardy scale, where fast but natural and ventilator deaths make it difficult to determine disease. Otherwise, controls are violent/fast deaths and "disease" is individuals who were previously ill. Finally, `recount_brain_v2` only uses public data and more in-depth information is private.


```{r}
#developmental stage
development <- "adult"

# DTHHRDY explanation
stat <- c("ventilator", "violent_fast", "fast_natural", "ill_unexpected", "ill_expected")

#GTEx disease mapping
gtex_disease <- c()
gtex_disease_status <- c()
for(i in GTEx_brain_merge$DTHHRDY) {
  gtex_disease <- c(gtex_disease, stat[i+1])
  if(i == 1) {
    gtex_disease_status <- c(gtex_disease_status, "Control")
  }
  if(i == 2 | i == 0) {
    gtex_disease_status <- c(gtex_disease_status, "either")
  }
  if(i == 3 | i == 4) {
    gtex_disease_status <- c(gtex_disease_status, "Disease")
  }
} 

# mapping sex to character
sex_character <- c("male", "female")
sex <- c()
for(i in GTEx_brain_merge$SEX) sex <- c(sex, sex_character[i])

#RNA isolation type
technique <- paste0("RNA Seq, ", GTEx_brain_merge$smnabtcht)

gtex_sampleInfo_brain_merged <- merge(GTEx_brain_merge, gtexSampinfo,
    by.x = "sampid", by.y = "SAMPID", all = F)
rownames(gtex_sampleInfo_brain_merged) <- gtex_sampleInfo_brain_merged$sampid
rownames(gtexSampinfo) <- gtexSampinfo$SAMPID
```

## Sample isolation

Sample isolation, fixed and frozen samples. These data were originally acquired from the pheno_sampid column. Time after isolation was taken from the `SMTSISCH` function and is currently being used as a rough proxy for Post-mortem interval. Afterwards, samples with `SMTSPAX > 0` were fixed, `SMTSISH < 0` were frozen. Otherwise it's tough to tell.

```{r}

gtexSampinfo$SMTSPAX[is.na(gtexSampinfo$SMTSPAX)] <- 0
gtexSampinfo$SMTSISCH[is.na(gtexSampinfo$SMTSISCH)] <- 0

isoTime <- rep(NA, nrow(GTEx_brain_merge))
names(isoTime) <- GTEx_brain_merge$sampid
count <- 1
for(i in names(isoTime)) {
  #print(which(rownames(gtexSampinfo) == i))
  isoTime[i] <- gtexSampinfo[i, "SMTSISCH"]
  
}

prep <- c()
for(i in 1:nrow(gtexSampinfo)) {
  if(gtexSampinfo$SMTSPAX[i] > 0) {
    prep[i] <- "fixed"
    next
  }
  if(gtexSampinfo$SMTSISCH[i] < 0) {
    prep[i] <- "frozen"
    next
  }
  prep[i] <- "unclear"
}
names(prep) <- rownames(gtexSampinfo)

prep_use <- rep(0, nrow(GTEx_brain_merge)) 
names(prep_use) <- GTEx_brain_merge$sampid


for(i in names(prep_use)){
  prep_use[i] <- prep[i]
}
```

## Combine

Combine metadata in the order of `recount_brain_v1`. Some columns (e.g. age units, sample location, public availability...) were consistent across all samples, are these columns are a consistent character vector.

```{r}
GTEX_combn <- cbind(GTEx_brain_merge$AGE, "Years", technique, GTEx_brain_merge$avg_read_length,
    NA, NA, "Laboratory, Data Analysis and Coordinating Center (LDACC)",
    broadman_gtex, NA, 
    "Laboratory, Data Analysis and Coordinating Center (LDACC)", NA,
    NA, "Public", "Adult", gtex_disease, gtex_disease_status, 
    GTEx_brain_merge$experiment,
    hemisphere_gtex, NA, "Illumina TruSeq RNA sequencing", NA, "paired",
    "cDNA", paste0("transcriptomic - ", GTEx_brain_merge$smcenter),
    GTEx_brain_merge$smnabtchd, NA,  NA, "Homo sapiens", NA, "Illumina",
    unlist(isoTime),
    "mins", unlist(prep_use), TRUE, "not public", GTEx_brain_merge$smnabtch,
    GTEx_brain_merge$smrin,
    GTEx_brain_merge$run, NA, GTEx_brain_merge$smts, sex,
    GTEx_brain_merge$sample, GTEx_brain_merge$project, tissue_1_gtex,
    NA, NA, NA, "Postmortem")


```

# TCGA

The code below makes the adjustments to TCGA. This code chunk extracts the brain (i.e. Lower Grade Glioma and Glioblastoma) samples from TCGA.

```{r}
# filter for brain samples 
tcga_brain_nums <- which(tcga$gdc_cases.project.project_id %in%
    c("TCGA-LGG", "TCGA-GBM"))
cd_brain.ol <- tcga[tcga_brain_nums,]
```

RNAseq file information. Average read length was calculated using the formula below:

$$ avgReadLength = auc / (mappedReadCount * numberEnds) $$

I.e. if RNA-seq was paired end the average read length was halved from

$$ auc/mappedReadCount$$

File size (mega bytes) is file size

$$ bytes / 1,000,000 $$


```{r}
#Avg read length
TCGA_readlength <- cd_brain.ol$auc / cd_brain.ol$mapped_read_count * ifelse(cd_brain.ol$paired_end, 2, 1)
#file size in megabytes
mb_tcga <- cd_brain.ol$gdc_file_size / 1e6

```

## Age and disease information

Age at diagnosis is used for age instead of age of treatment/death. These information can still be acquired from the TCGA metadata information. merge these data with `toupper(cd_brain.ol$gdc_file_id)`, which is the identifier that maps to the row names of the TCGA count data. The youngest individual in the TCGA_brain dataset is 14, so for development samples are split into adolescent/adult. 

Disease is information from `gdc_cases.samples.sample_type`. disease status is binary from if the tissue was disease or normal tissues (5 samples). Also tumour-cDNA or cDNA is split the same way.



```{r}
#age at diagnosis
age_at_diag <- cd_brain.ol$cgc_case_age_at_diagnosis

#age normalized for developmental stage
development_tcga <- c()
for(i in 1:length(cd_brain.ol$cgc_case_age_at_diagnosis)) {
  if(is.na(age_at_diag[i])) {
    development_tcga[i] <- NA
    next
  } 
  if(age_at_diag[i] < 20) {
    development_tcga[i] <- "Adolescent"
    next
  }
  development_tcga[i] <- "Adult"
}

#cDNA type and solid tissue normal
disease_status_tcga <- c()
selection_tcga <- c()
for(i in 1:length(cd_brain.ol$gdc_cases.samples.sample_type)) {
  if(cd_brain.ol$gdc_cases.samples.sample_type[i] ==  "Solid Tissue Normal") {
    disease_status_tcga[i] <- "Control"
    selection_tcga[i] <- "cDNA"
  } else {
    disease_status_tcga[i] <- "Disease"
    selection_tcga[i] <- "ctDNA"
  }
}

# Histological grade, data is changed to match recount brain
neoP <- tochr(cd_brain.ol$xml_neoplasm_histologic_grade)
neoP[is.na(neoP)] <- "0"
grade_adjust <- c()
for(i in 1:length(neoP)) {
  if(neoP[i] ==  "G2") {
    grade_adjust[i] <- "Grade II"
    next
  }
  if(neoP[i] == "G3") {
    grade_adjust[i] <- "Grade III"
    next
  }
  grade_adjust[i] <- NA
}

path <- cd_brain.ol$xml_ldh1_mutation_found
pathology_comp <- c()
for(i in 1:length(path)) {
  if(is.na(path[i])) {
    pathology_comp[i] <- NA
    next
  }
  if(path[i] == "YES") {
    pathology_comp[i] <- "+ IDH1 Mutation"
    next
  }
  if(path[i] == "NO") {
    pathology_comp[i] <- "- IDH1 Mutation"
    next
  }
  pathology_comp[i] <- path[i]
}
table(pathology_comp)

# LGG or GBM
cancer_type <- substr(x = cd_brain.ol$gdc_cases.project.project_id, 6,
    nchar(cd_brain.ol$gdc_cases.project.project_id))
```

## Combine

Combining TCGA data into the recount_brain_v1 format, some columns are consistent (e.g. all sequencing data was paired end) As such, the paired end sequencing column is "paired".

```{r}
TCGA_combn <- cbind(cd_brain.ol$cgc_case_age_at_diagnosis, "Years", "RNA_seq", TCGA_readlength,
    cd_brain.ol$cgc_case_id, cd_brain.ol$xml_patient_id,
    cd_brain.ol$gdc_cases.tissue_source_site.name,
    NA, NA, cd_brain.ol$gdc_center.name, grade_adjust,
    cd_brain.ol$gdc_cases.samples.sample_type,
    cd_brain.ol$gdc_metadata_files.access.analysis, development_tcga,
    cd_brain.ol$gdc_cases.samples.sample_type, disease_status_tcga,
    cd_brain.ol$gdc_metadata_files.file_id.experiment, NA, NA,
    cd_brain.ol$gdc_platform, toupper(cd_brain.ol$gdc_file_id), "paired", 
    selection_tcga, "TRANSCRIPTOMIC", cd_brain.ol$cgc_file_upload_date, NA,
    cd_brain.ol$gdc_file_size / 1e6, "Homo sapiens", pathology_comp,
    "Illumina", NA, NA, "frozen soon after surgery", "TRUE",
    cd_brain.ol$gdc_cases.demographic.race,
    cd_brain.ol$cgc_file_published_date, NA, NA, NA,"Brain",
    cd_brain.ol$gdc_cases.demographic.gender, NA,NA,cancer_type,NA,NA,
    tochr(cd_brain.ol$xml_histological_type), "Biopsy")
                    
rownames(cd_brain.ol) <- toupper(cd_brain.ol$gdc_file_id)

```

## Drug information

The drug information in the `cgc_drug_therapy_drug_name` column contains multiple typos and ambiguous drug names. The script below adjusts these drug names to allow for consistency. `drug_info_T` informs the presence of drug information. `drug_therapy_type` distinguishes between chemo, radiation etc. Finally, 260/280 is the TCGA proxy of RNA quality. Some older cancers (i.e. OV) have RIN, however LGG and GBM moved over to 260/280. 

```{r}
dN <- toupper(cd_brain.ol$cgc_drug_therapy_drug_name)
drugName <- c()
# fixed typos in TCGA drugs
for(i in 1:length(dN)) { 
  if(is.na(dN[i])) {
    drugName[i] <- NA
    next
  }
  if(dN[i] %in% c("TEMOZOLAMIDE", "TEMOZOLOMIDE")) {
    drugName[i] <- "TEMOZOLOMIDE"
    next
  }
  
  if(dN[i] %in% c("TEMADOR","TEMODAR", "TEMODAR (ESCALATION)", "METRONOMIC TEMODAR")) {
    drugName[i] <- "TEMODAR"
    next
  }
  
  if(dN[i] %in% c("LOMUSTINE (CCNU)","LOMUSTINE", "LOMUSTIN")) {
    drugName[i] <- "LOMUSTINE"
    next
  }
  
  if(dN[i] %in% c("ISOTRETINOIN","ISOTRECTINOIN (ACCCUTANE)")) {
    drugName[i] <- "ISOTRETINOIN"
    next
  }
  
  if(dN[i] %in% c("I 131 81C6","I131-81C6")) {
    drugName[i] <- "I-131-81C6"
    next
  }
  
  if(dN[i] %in% c("HYDROXYUREA","HYDROYUREA")) {
    drugName[i] <- "HYDROXYUREA"
    next
  }
  
  if(dN[i] %in% c("GLIADEL WAFER","GLIADEL WAFER (BCNU)", "GLIADEL")) {
    drugName[i] <- "GLIADEL"
    next
  }
  
  if(dN[i] %in% c("DEXAMETHASONE","DEXMETHASONE")) {
    drugName[i] <- "DEXAMETHASONE"
    next
  }
  
  if(dN[i] %in% c("CPT11","CPT-11")) {
    drugName[i] <- "CPT11"
    next
  }
  
  if(dN[i] %in% c("CARMUSTINE", "CARMUSTIN", "CARMUSTINE (BCNU)", "CARMUSTINE BCNU")) {
    drugName[i] <- "CARMUSTINE"
    next
  }
  
  if(dN[i] %in% c("BEVACIZUMAB","BEVACIZUMAB OR PLACEBO RTOG 0825")) {
    drugName[i] <- "BEVACIZUMAB"
    next
  }
  
  if(dN[i] %in% c("BCNU","BCNU (CARMUSTINE)")) {
    drugName[i] <- "BCNU"
    next
  }
  drugName[i] <- dN[i]
}

drug_info_T <- cd_brain.ol$xml_has_drugs_information
drug_therapy_type <- cd_brain.ol$cgc_drug_therapy_pharmaceutical_therapy_type

T_260_280 <- cd_brain.ol$gdc_cases.samples.portions.analytes.a260_a280_ratio
```

# Combining all and cleaning

The code below readjusts the order of `recount_brain_v1`. This is completed by insuring that the order of columns in recount_brain matches TCGA, GTEx, and the recount website. All of the column names are then matches.

```{r}
notes48 <- notes[1:48,]
recount_brain_reorder <- recount_brain[,notes48$recount_brain]

colnames(TCGA_combn) <- colnames(GTEX_combn) <- colnames(recount_brain_reorder)
```


The data below cleans up the `colData` related to combining the three datasets and making a consistent identifier. The `Study` is the name of SRA study, `TCGA`, or `GTEX`. The `_full` columns are TCGA columns with the correct number of rows filled up for recount_brain and GTEx. Finally, these columns are combined together and the dataset is saved.

```{r}
Study <- sub("\\..*","", rownames(recount_brain) )
Study_full <- c(Study, rep("TCGA", nrow(TCGA_combn)),
    rep("GTEX", nrow(GTEX_combn)))
Dataset <- c(rep("recount_brain_v1",length(Study)),
    rep("TCGA", nrow(TCGA_combn)), rep("GTEX", nrow(GTEX_combn)))
drugName_full <- c(rep(NA, length(Study)), drugName, rep(NA, nrow(GTEX_combn) ))
drug_info_full <- c(rep(NA, length(Study)), drug_info_T,
    rep(NA, nrow(GTEX_combn) ))
drug_type_full <- c(rep(NA, length(Study)), drug_therapy_type,
    rep(NA, nrow(GTEX_combn) ))
full_260_280<- c(rep(NA, length(Study)), T_260_280, rep(NA, nrow(GTEX_combn) ))
count_file_identifier <- c(recount_brain$run_s, rownames(cd_brain.ol),
    GTEx_brain_merge$run)
brain_meta <- rbind(recount_brain_reorder, TCGA_combn, GTEX_combn )
metadata_complete <- cbind(brain_meta, Study_full, drugName_full,
    drug_info_full, drug_type_full, full_260_280, count_file_identifier, Dataset)
```

## Consistent names

The code below adjusts some of the major columns within the dataset to account for different datasets using slightly different names. For example, if you filter for "Primary", you get all primary tumors instead of just the recount_brain_v1 primary tumors.

```{r}

#Tissue site 1 adjust

tsite1 <- c()
ts <- metadata_complete$tissue_site_1
for(i in 1:nrow(metadata_complete)) {
  
  if(ts[i] %in% c("Caudate (basal ganglia)", "Caudate")) {
    tsite1[i] <- "Caudate"
    next
  }
  
  if(ts[i] %in% c("Frontal Cortex", "Frontal Cortex (BA9)")) {
    tsite1[i] <- "Frontal Cortex"
    next
  }
  
  if(ts[i] %in% c("Nucleus accumbens", "Nucleus accumbens (basal ganglia)")) {
    tsite1[i] <- "Nucleus accumbens"
    next
  }
  
  if(ts[i] %in% c("Putamen", "Putamen (basal ganglia)")) {
    tsite1[i] <- "Putamen"
    next
  }
 tsite1[i] <- ts[i] 
}

metadata_complete$tissue_site_1 <- tsite1


# Adjusting the disease information so that tumour information is consistent 

dis <- c() # Note, In azheimer's disease and Parkinson's disease there was a minor error with the encoding of the apostrophe. You will likely need to adjust these individuals manually 
for(i in 1:length(metadata_complete$disease)) {
  if(metadata_complete$disease[i] %in% c("Brain tumor", "Tumor")) {
    dis[i] <- "brain tumor unspecified"
    next
  }
  dis[i] <- metadata_complete$disease[i]
}

metadata_complete$disease <- dis

clinStage2 <- c()
for(i in 1:length(metadata_complete$clinical_stage_2)) {
  if(is.na(metadata_complete$clinical_stage_2[i])) {
    clinStage2[i] <- NA
    next
  }
  if(metadata_complete$clinical_stage_2[i] %in% c("Primary Tumor")) {
    clinStage2[i] <- "Primary"
    next
  }
  if(metadata_complete$clinical_stage_2[i] %in% c("Recurrent Tumor")) {
    clinStage2[i] <- "Recurrent"
    next
  }
  clinStage2[i] <- metadata_complete$clinical_stage_2[i]
}

metadata_complete$clinical_stage_2 <- clinStage2

# Fixing capital in consernt
metadata_complete$consent_s <- toupper(metadata_complete$consent_s)

race_adjusted <- toupper(metadata_complete$race)
for(i in 1:length(race_adjusted)) {
  if(race_adjusted[i] %in% "BLACK OR AFRICAN AMERICAN") {
    race_adjusted[i] <- "BLACK"
  }  
  
}
metadata_complete$race <- race_adjusted


#Information on sample origin: iPSC conistency
origin <- metadata_complete$sample_origin
for(i in 1:length(origin)) {
    if(origin[i] %in% "iPSCs") {
    origin[i] <- "iPSC"
  }  
}
metadata_complete$sample_origin <- origin

# making sure that oligodendroglioma/oligodendrogliomas are different
t_type <- metadata_complete$tumor_type
for(i in 1:length(t_type)) {
  if(t_type[i] %in% "Anaplastic Oligodendrogliomas") {
    t_type[i] <- "Anaplastic Oligodendroglioma"
    next
  }  
  

}
metadata_complete$sample_origin <- origin
metadata_complete$tumor_type <- t_type

# Converting run_s to also contain the identifier. This allows recount_brain_v2 to be accessed via the "add_metadata()" function

metadata_complete$run_s <-  metadata_complete$count_file_identifier

```

## Create `recount_brain_v2`

The final code chunk checks the final dimensions and md5sum object of `recount_brain_v2` before saving it into an Rdata object and listing variables.

```{r}
#Completed metadata is the combined and saved
recount_brain <- metadata_complete
dim(recount_brain)

## For compatibility with add_metadata()
recount_brain$run_s <- as.character(recount_brain$run_s)

## Re-cast some vars
recount_brain$count_file_identifier <- as.character(recount_brain$count_file_identifier)
recount_brain$drug_info_full <- recount_brain$drug_info_full == 'YES'
recount_brain$rin <- as.numeric(recount_brain$rin)
recount_brain$pmi <- as.numeric(recount_brain$pmi)
recount_brain$avgspotlen_l <- as.numeric(recount_brain$avgspotlen_l)
recount_brain$insertsize_l <- as.numeric(recount_brain$insertsize_l)
recount_brain$mbases_l <- as.integer(recount_brain$mbases_l)
recount_brain$mbytes_l <- as.numeric(recount_brain$mbytes_l)
recount_brain$brodmann_area <- as.integer(recount_brain$brodmann_area)
recount_brain$present_in_recount <- as.logical(recount_brain$present_in_recount)

## Simplify age by turning ranges such as 20-29 to mean(c(20, 29))
mean_age <- function(x) {
    mean(as.integer(strsplit(x, '-')[[1]]))
}
age <- as.numeric(recount_brain$age)
age[grepl('-', recount_brain$age)] <- sapply(
    recount_brain$age[grepl('-', recount_brain$age)], mean_age)
recount_brain$age <- age

## Between version 1 and 2, these are the columns that change types
r <- add_metadata(source = 'recount_brain_v1')
x <- sapply(r, class) == sapply(recount_brain[, colnames(r)], class)
sapply(recount_brain[, colnames(r)], class)[!x]
sapply(r, class)[!x]

## Save the data
save(recount_brain, file = 'recount_brain_v2_noOntology.Rdata')
write.csv(recount_brain, file = 'recount_brain_v2_noOntology.csv', quote = TRUE,
    row.names = FALSE)

## Check md5sum for the resulting files
sapply(dir(pattern = 'recount_brain_v2'), tools::md5sum)

## List of all variables
colnames(recount_brain)
```

# Explore `recount_brain_v2`

Below provides some summary statistics on the merged dataset. Below there are some pivot tables of columns split by the major dataset.

```{r}
#Sex
table(recount_brain$sex, recount_brain$Dataset) 

#Development 
table(recount_brain$development, recount_brain$Dataset) 

#Tumor type
table(recount_brain$tumor_type, recount_brain$Dataset) 

# Clinical stage 2
table(recount_brain$clinical_stage_2, recount_brain$Dataset) 

# tissue_site 1
table(recount_brain$tissue_site_1, recount_brain$Dataset) 

# present in recount

table(recount_brain$present_in_recount, recount_brain$Dataset) 

```

Full summary:

```{r}
summary(recount_brain)
```



# Reproducibility

This document was made possible thanks to:

* R `r citep(bib[['R']])`
* `r Biocpkg('BiocStyle')` `r citep(bib[['BiocStyle']])`
* `r CRANpkg('devtools')` `r citep(bib[['devtools']])`
* `r CRANpkg('knitcitations')` `r citep(bib[['knitcitations']])`
* `r CRANpkg('knitr')` `r citep(bib[['knitr']])`
* `r Biocpkg('recount')` `r citep(bib[['recount']])`
* `r CRANpkg('rmarkdown')` `r citep(bib[['rmarkdown']])`

Code for creating this document
 
```{r createVignette, eval=FALSE}
## Create the vignette
library('rmarkdown')
system.time(render('cross_studies_metadata.Rmd', 'BiocStyle::html_document'))
```


Reproducibility information for this document.

```{r 'reproducibility info'}
## Reproducibility info
proc.time()
message(Sys.time())
options(width = 120)
session_info()
```

# Bibliography

This document was generated using `r Biocpkg('BiocStyle')` `r citep(bib[['BiocStyle']])` with `r CRANpkg('knitr')` `r citep(bib[['knitr']])` and `r CRANpkg('rmarkdown')` `r citep(bib[['rmarkdown']])` running behind the scenes.

Citations made with `r CRANpkg('knitcitations')` `r citep(bib[['knitcitations']])` and the bibliographical file is available [here](cross_studies_metadata.bib).

```{r biblio, results = 'asis', echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
bibliography()
```
