---
title: "Ontology-based metadata for recount-brain"
author: "Sean Davis"
date: "12/11/2018"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Build ontology lookup

```{r}
library(tidyverse)
build_ontology_table = function(ontology_name) {
    destfile = paste0(ontology_name,".csv.gz")
    download.file(sprintf("http://data.bioontology.org/ontologies/%s/download?apikey=8b5b7825-538d-40e0-9e9e-5ab9274a9aeb&download_format=csv", ontology_name),
                  destfile = destfile)
    tmptab = read.csv(destfile, header = TRUE, stringsAsFactors = FALSE)[,1:8]
    tmptab$ontology = ontology_name
    unlink(destfile)
    return(as_tibble(tmptab))
}
```

```{r message=FALSE}
ontolookup = bind_rows(build_ontology_table('ORDO'),
                       build_ontology_table('HP'),
                       build_ontology_table('UBERON'),
                       build_ontology_table('DOID'),
                       build_ontology_table('NCIT'))
```

```{r}
load("recount_brain_v2.Rdata")
head(recount_brain)
```

## Brodmann area

The UBERON ontology has direct mappings to Brodmann areas, amenable to direct lookup.

```{r}
extended_brodmann_area = paste("Brodmann (1909) area", recount_brain$brodmann_area)
ontomap = ontolookup$`Class.ID`[match(unique(extended_brodmann_area), ontolookup$Preferred.Label )]
names(ontomap) = unique(extended_brodmann_area)
ontomap['Brodmann (1909) area 41'] = "http://purl.obolibrary.org/obo/UBERON_0006095"
matchers = match( extended_brodmann_area, names(ontomap))
ontobrain = data.frame(brodmann_ontology = ontomap[matchers])
```

## Disease

```{r}
disease_mappings = c(
    "Bipolar disorder" = "http://purl.obolibrary.org/obo/HP_0007302",
    "Schizophrenia" = "http://purl.obolibrary.org/obo/HP_0100753",
    "Hutchinson-Gilford progeria syndrome" = "http://www.orpha.net/ORDO/Orphanet_740",
    "Cortical ischemic stroke tissue" = "http://purl.obolibrary.org/obo/HP_0002140",
    "Autism spectrum disorder" = "http://purl.obolibrary.org/obo/HP_0000717",
    "Epilepsy" = "http://purl.obolibrary.org/obo/DOID_1826",
    "brain tumor unspecified" = "http://purl.obolibrary.org/obo/HP_0030692",
    "Huntington's disease" = "http://www.orpha.net/ORDO/Orphanet_399",
    "Alzheimer’s disease" = "http://purl.obolibrary.org/obo/HP_0002511",
    "Parkinson’s disease" = "http://purl.obolibrary.org/obo/DOID_14332",
    "Parkinson's disease" = "http://purl.obolibrary.org/obo/DOID_14332",
    "Rett syndrome" = "http://www.orpha.net/ORDO/Orphanet_778",
    "Spinal muscular atrophy" = "http://purl.obolibrary.org/obo/HP_0007269",
    "Amyotrophic lateral sclerosis" = "http://www.orpha.net/ORDO/Orphanet_803",
    "Angelman syndrome" = "http://www.orpha.net/ORDO/Orphanet_72",
    "Embryonal tumors with multilayered rosettes" = "http://ncicb.nci.nih.gov/xml/owl/EVS/Thesaurus.owl#C4915",
    "Primitive neuroectodermal tumor" = "http://ncicb.nci.nih.gov/xml/owl/EVS/Thesaurus.owl#C3716",
    "Primary Tumor" = "http://purl.obolibrary.org/obo/HP_0030692",
    "Recurrent Tumor" = "http://ncicb.nci.nih.gov/xml/owl/EVS/Thesaurus.owl#C153837"
)
matchers = match( recount_brain$disease, names(disease_mappings))
ontobrain$disease_ontology = disease_mappings[matchers]
ontomap = c(ontomap, disease_mappings)
```

## Tissue site

Since ontologies are graphs, we only need to annotate the most specific term. 


```{r}
tissues = recount_brain$tissue_site_1
not_na = !is.na(recount_brain$tissue_site_2)
tissues[not_na] = recount_brain$tissue_site_2[not_na]
not_na = !is.na(recount_brain$tissue_site_3)
tissues[not_na] = recount_brain$tissue_site_3[not_na]
# use only uberon
uberon = ontolookup[ontolookup$ontology=='UBERON',]
matcher = function(term) {
    tmp = grep(paste0('^', term, '$'), uberon$Preferred.Label, ignore.case = TRUE)
    if(length(tmp)==1)  {
        tmp = uberon$Class.ID[tmp]
        names(tmp) = term
        return(tmp)
    } else {
        tmp = NA
        names(tmp) = term
        return(tmp)
    }
}
tissue_mappings = sapply(unique(tissues), matcher, USE.NAMES = FALSE)
names(tissue_mappings)[is.na(tissue_mappings)]
tissue_mappings = c(tissue_mappings, 
                    c(
    "Motor cortex" = "http://purl.obolibrary.org/obo/UBERON_0001384",
    "Anterior cingulate gyrus" = "http://purl.obolibrary.org/obo/UBERON_0002756",
    "Anterior prefrontal cortex" = "http://purl.obolibrary.org/obo/UBERON_0000451",
    "Whole brain" = "http://purl.obolibrary.org/obo/UBERON_0000955",
    "Mixed" = "http://purl.obolibrary.org/obo/UBERON_0000955",
    "Hippocampus" = "http://purl.obolibrary.org/obo/UBERON_0002421",
    "Anterior cingulate cortex (BA24)" = "http://purl.obolibrary.org/obo/UBERON_0006101",
    "Spinal cord (cervical c-1)" = "http://purl.obolibrary.org/obo/UBERON_0002726",
    "GBM" = "http://purl.obolibrary.org/obo/HP_0012174",
    "LGG" = "http://ncicb.nci.nih.gov/xml/owl/EVS/Thesaurus.owl#C132067"
))
ontomap = c(ontomap, tissue_mappings)
matchers = match(tissues, names(tissue_mappings))
ontobrain$tissue = tissues
ontobrain$tissue_ontology = tissue_mappings[matchers]
```

## Result files

The `recount_brain_v2_with_ontology.csv` file contains the 
metadata with ontologies for Tissue, Brodmann area, and Disease added.

The `ontology_mapping.csv` file contains a dataframe of the mappings
from original names to ontology entries.

```{r}
recount_brain_with_ontology = cbind(recount_brain, ontobrain)
write.csv(recount_brain_with_ontology, file="recount_brain_v2_with_ontology.csv")
ontomap_full = merge(as.data.frame(ontomap), ontolookup, by.x=1, by.y = 1, all.x=TRUE)
ontomap_full$original_name = names(ontomap)
write.csv(ontomap_full, file='ontology_mapping.csv')
```
