---
title: "Ontology-based metadata for recount-brain"
author: 
  - name: Sean Davis
    affiliation:
    - Genetics Branch, Center for Cancer Research, National Cancer Institute, NIH, Bethesda, MD, 20892, USA
    email: sdavis2@mail.nih.gov
  - name: Leonardo Collado-Torres
    affiliation:
    - &libd Lieber Institute for Brain Development, Johns Hopkins Medical Campus
    - &ccb Center for Computational Biology, Johns Hopkins University
    email: leo.collado@libd.org
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This code takes the `recount-brain_v2` output created by Dustin Sokolowski in [`cross_studies_metadata`](cross_studies_metadata.html) and adds ontology information. Specifically, this adds the `brodmann_ontology`, `disease_ontology`, `tissue` and `tissy_ontology` fields to `recount-brain`. The results are saved in [`recount_brain_v2.Rdata`](recount_brain_v2.Rdata) (as well as a csv version) which can then be used with `recount::add_metadata(source = 'recount_brain_v2')`.


```{r 'citation info', warning = FALSE, message = FALSE, echo = FALSE}
library('knitcitations')
library('BiocStyle')


## Load knitcitations with a clean bibliography
cleanbib()
cite_options(hyperlink = 'to.doc', citation_format = 'text', style = 'html')
# Note links won't show for now due to the following issue
# https://github.com/cboettig/knitcitations/issues/63
 ## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation('BiocStyle'),
    bioportal = bib_metadata('10.1093/nar/gkr469'),
    sessioninfo = citation('sessioninfo'),
    knitcitations = citation('knitcitations'),
    knitr = citation('knitr')[3],
	tidyverse = citation('tidyverse'),
    rmarkdown = citation('rmarkdown')
)
write.bibtex(bib, file = 'recount_brain_ontologies.bib')
```


# Build ontology lookup

First we start by building an ontology lookup table using data from [data.bioontology.org/](http://data.bioontology.org/)'s API.

```{r 'build onto table code', message = FALSE}
library('tidyverse')
build_ontology_table = function(ontology_name) {
    destfile = paste0(ontology_name,".csv.gz")
    download.file(sprintf("http://data.bioontology.org/ontologies/%s/download?apikey=8b5b7825-538d-40e0-9e9e-5ab9274a9aeb&download_format=csv", ontology_name),
                  destfile = destfile)
    tmptab = read.csv(destfile, header = TRUE, stringsAsFactors = FALSE)[,1:8]
    tmptab$ontology = ontology_name
    unlink(destfile)
    return(as_tibble(tmptab))
}
```

```{r 'build onto table', message=FALSE}
ontolookup = bind_rows(build_ontology_table('ORDO'),
                       build_ontology_table('HP'),
                       build_ontology_table('UBERON'),
                       build_ontology_table('DOID'),
                       build_ontology_table('NCIT'))
```

Next we load the `recount-brain` files created by Dustin Sokolowski and explore them.

```{r 'load recount brain'}
load("recount_brain_v2_noOntology.Rdata")
head(recount_brain)
```

## Brodmann area

The UBERON ontology has direct mappings to Brodmann areas, amenable to direct lookup.

```{r 'find the brodmann area'}
extended_brodmann_area = paste("Brodmann (1909) area", recount_brain$brodmann_area)
ontomap = ontolookup$`Class.ID`[match(unique(extended_brodmann_area), ontolookup$Preferred.Label )]
names(ontomap) = unique(extended_brodmann_area)
ontomap['Brodmann (1909) area 41'] = "http://purl.obolibrary.org/obo/UBERON_0006095"
matchers = match( extended_brodmann_area, names(ontomap))
ontobrain = data.frame(
    brodmann_ontology = ontomap[matchers],
    stringsAsFactors = FALSE
)

## Add synonyms too, for searching by text
ontobrain$brodmann_synonyms <- ontolookup$Synonyms[
    match(ontobrain$brodmann_ontology, ontolookup$Class.ID)]
```

```{r 'brodmann area parent info'}
## Functions for traversing the ontolookup table to find all the parents
find_parent <- function(i) {
    res <- ontolookup$Parents[ontolookup$Class.ID == i]
    if(length(res) == 0) return(NULL)
    strsplit(res, '\\|')[[1]]
}
find_parent_recurr <- function(i) {
    res <- find_parent(i)
    if(length(res) == 0) {
        return(NULL)
    } else {
        return(c(res, unique(unlist(sapply(res, find_parent_recurr)))))
    }
}

## Find the unique brodmann ontologies listed
brodmann_uni <- as.character(unique(ontobrain$brodmann_ontology))
brodmann_uni <- brodmann_uni[!is.na(brodmann_uni)]

## Find the parents
brodmann_parents <- lapply(brodmann_uni, find_parent_recurr)
names(brodmann_parents) <- brodmann_uni

## Find what they mean
brodmann_parents_label <- lapply(brodmann_parents, function(x) {
    ontolookup$Preferred.Label[match(x, ontolookup$Class.ID)]
})

## Could have also found the parents synonyms, but it doesn't
## add much
# lapply(brodmann_parents, function(x) ontolookup$Synonyms[match(x, ontolookup$Class.ID)])

## Add them and their labels
ontobrain$brodmann_parents <- sapply(brodmann_parents, paste, collapse = ' | ')[
    match(ontobrain$brodmann_ontology, brodmann_uni)]
ontobrain$brodmann_parents_label <- sapply(brodmann_parents_label, paste, collapse = ' | ')[
    match(ontobrain$brodmann_ontology, brodmann_uni)]
```


## Disease

We then select the disease ontology mappings.

```{r 'map diseases'}
disease_mappings = c(
    "Bipolar disorder" = "http://purl.obolibrary.org/obo/HP_0007302",
    "Schizophrenia" = "http://purl.obolibrary.org/obo/HP_0100753",
    "Hutchinson-Gilford progeria syndrome" = "http://www.orpha.net/ORDO/Orphanet_740",
    "Cortical ischemic stroke tissue" = "http://purl.obolibrary.org/obo/HP_0002140",
    "Autism spectrum disorder" = "http://purl.obolibrary.org/obo/HP_0000717",
    "Epilepsy" = "http://purl.obolibrary.org/obo/DOID_1826",
    "brain tumor unspecified" = "http://purl.obolibrary.org/obo/HP_0030692",
    "Huntington's disease" = "http://www.orpha.net/ORDO/Orphanet_399",
    "Alzheimer’s disease" = "http://purl.obolibrary.org/obo/HP_0002511",
    "Parkinson’s disease" = "http://purl.obolibrary.org/obo/DOID_14332",
    "Parkinson's disease" = "http://purl.obolibrary.org/obo/DOID_14332",
    "Rett syndrome" = "http://www.orpha.net/ORDO/Orphanet_778",
    "Spinal muscular atrophy" = "http://purl.obolibrary.org/obo/HP_0007269",
    "Amyotrophic lateral sclerosis" = "http://www.orpha.net/ORDO/Orphanet_803",
    "Angelman syndrome" = "http://www.orpha.net/ORDO/Orphanet_72",
    "Embryonal tumors with multilayered rosettes" = "http://ncicb.nci.nih.gov/xml/owl/EVS/Thesaurus.owl#C4915",
    "Primitive neuroectodermal tumor" = "http://ncicb.nci.nih.gov/xml/owl/EVS/Thesaurus.owl#C3716",
    "Primary Tumor" = "http://purl.obolibrary.org/obo/HP_0030692",
    "Recurrent Tumor" = "http://ncicb.nci.nih.gov/xml/owl/EVS/Thesaurus.owl#C153837"
)
matchers = match( recount_brain$disease, names(disease_mappings))
ontobrain$disease_ontology = disease_mappings[matchers]
ontomap = c(ontomap, disease_mappings)
```

## Tissue site

Since ontologies are graphs, we only need to annotate the most specific term. 


```{r 'map tissues'}
tissues = recount_brain$tissue_site_1
not_na = !is.na(recount_brain$tissue_site_2)
tissues[not_na] = recount_brain$tissue_site_2[not_na]
not_na = !is.na(recount_brain$tissue_site_3)
tissues[not_na] = recount_brain$tissue_site_3[not_na]
# use only uberon
uberon = ontolookup[ontolookup$ontology=='UBERON',]
matcher = function(term) {
    tmp = grep(paste0('^', term, '$'), uberon$Preferred.Label, ignore.case = TRUE)
    if(length(tmp)==1)  {
        tmp = uberon$Class.ID[tmp]
        names(tmp) = term
        return(tmp)
    } else {
        tmp = NA
        names(tmp) = term
        return(tmp)
    }
}
tissue_mappings = sapply(unique(tissues), matcher, USE.NAMES = FALSE)
names(tissue_mappings)[is.na(tissue_mappings)]
tissue_mappings = c(tissue_mappings, 
                    
    "Motor cortex" = "http://purl.obolibrary.org/obo/UBERON_0001384",
    "Anterior cingulate gyrus" = "http://purl.obolibrary.org/obo/UBERON_0002756",
    "Anterior prefrontal cortex" = "http://purl.obolibrary.org/obo/UBERON_0000451",
    "Whole brain" = "http://purl.obolibrary.org/obo/UBERON_0000955",
    "Mixed" = "http://purl.obolibrary.org/obo/UBERON_0000955",
    "Hippocampus" = "http://purl.obolibrary.org/obo/UBERON_0002421",
    "Anterior cingulate cortex (BA24)" = "http://purl.obolibrary.org/obo/UBERON_0006101",
    "Spinal cord (cervical c-1)" = "http://purl.obolibrary.org/obo/UBERON_0002726",
    "GBM" = "http://purl.obolibrary.org/obo/HP_0012174",
    "LGG" = "http://ncicb.nci.nih.gov/xml/owl/EVS/Thesaurus.owl#C132067"
)
ontomap = c(ontomap, tissue_mappings)
matchers = match(tissues, names(tissue_mappings))
ontobrain$tissue = tissues
ontobrain$tissue_ontology = tissue_mappings[matchers]

## Add synonyms too, for searching by text
ontobrain$tissue_synonyms <- ontolookup$Synonyms[
    match(ontobrain$tissue_ontology, ontolookup$Class.ID)]
```

```{r 'tissue parent info'}
## Find the unique tissue ontologies listed
tissue_uni <- as.character(unique(ontobrain$tissue_ontology))
tissue_uni <- tissue_uni[!is.na(tissue_uni)]

## Find the parents
tissue_parents <- lapply(tissue_uni, find_parent_recurr)
names(tissue_parents) <- tissue_uni

## Find what they mean
tissue_parents_label <- lapply(tissue_parents, function(x) {
    ontolookup$Preferred.Label[match(x, ontolookup$Class.ID)]
})

## Could have also found the parents synonyms, but it doesn't
## add much
# lapply(tissue_parents, function(x) ontolookup$Synonyms[match(x, ontolookup$Class.ID)])

## Add them and their labels
ontobrain$tissue_parents <- sapply(tissue_parents, paste, collapse = ' | ')[
    match(ontobrain$tissue_ontology, tissue_uni)]
ontobrain$tissue_parents_label <- sapply(tissue_parents_label, paste, collapse = ' | ')[
    match(ontobrain$tissue_ontology, tissue_uni)]
```

## Result files

The [`recount_brain_v2.Rdata`](recount_brain_v2.Rdata) and [`recount_brain_v2.csv`](recount_brain_v2.csv) files contain the metadata with ontologies for Tissue, Brodmann area, and Disease added. They also include the actual tissue as prioritized by `tissue_site_3` over `tissue_site_2` over `tissue_site_1`. For the Broadmann area and the tissue, we further added their synonyms, ontology parents and ontology parent labels which will facilitate text based searches on `recount_brain`.

The `ontology_mapping.csv` file contains a dataframe of the mappings
from original names to ontology entries.

```{r}
recount_brain = cbind(recount_brain, ontobrain)
write.csv(recount_brain, file="recount_brain_v2.csv", row.names = FALSE)

## Export in Rdata for use with recount::add_metadata()
save(recount_brain, file = 'recount_brain_v2.Rdata')
ontomap_full = merge(as.data.frame(ontomap), ontolookup, by.x=1, by.y = 1, all.x=TRUE)
ontomap_full$original_name = names(ontomap)
ontomap_full = ontomap_full[!is.na(ontomap_full$ontomap),]
write.csv(ontomap_full, file='ontology_mapping.csv')
```

Next we can explore which are the ontologies present in `recount_brain_v2`:

```{r}
map(
    recount_brain[, grep('ontology', colnames(recount_brain))],
    ~ sort(table('ontologies:' = .x), decreasing = TRUE)
)
```


Here's some information for making sure the files you download are complete.

```{r}
## Check md5sum for the resulting files
map(dir(pattern = 'recount_brain_v2|ontology_mapping'), tools::md5sum)

## List of all variables
colnames(recount_brain)
```


# Reproducibility

This document was made possible thanks to `BioPortal` `r citep(bib[['bioportal']])` and:

* R `r citep(bib[['R']])`
* `r Biocpkg('BiocStyle')` `r citep(bib[['BiocStyle']])`
* `r CRANpkg('knitcitations')` `r citep(bib[['knitcitations']])`
* `r CRANpkg('knitr')` `r citep(bib[['knitr']])`
* `r CRANpkg('rmarkdown')` `r citep(bib[['rmarkdown']])`
* `r CRANpkg('sessioninfo')` `r citep(bib[['sessioninfo']])`
* `r CRANpkg('tidyverse')` `r citep(bib[['tidyverse']])`

Code for creating this document
 
```{r createVignette, eval=FALSE}
## Create the vignette
library('rmarkdown')
system.time(render('recount_brain_ontologies.Rmd', 'BiocStyle::html_document'))
```


Reproducibility information for this document.

```{r 'reproducibility info'}
## Reproducibility info
proc.time()
message(Sys.time())
options(width = 120)
library('sessioninfo')
session_info()
```

# Bibliography

This document was generated using `r Biocpkg('BiocStyle')` `r citep(bib[['BiocStyle']])` with `r CRANpkg('knitr')` `r citep(bib[['knitr']])` and `r CRANpkg('rmarkdown')` `r citep(bib[['rmarkdown']])` running behind the scenes.

Citations made with `r CRANpkg('knitcitations')` `r citep(bib[['knitcitations']])` and the bibliographical file is available [here](recount_brain_ontologies.bib).

```{r biblio, results = 'asis', echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
bibliography()
```